---
title: "Introduction to linear regression"
author: "Christian Thieme"
date: "March 27, 2020"
output:
  pdf_document: default
  html_document:
    css: ./lab.css
    highlight: pygments
    theme: cerulean
---

## Batter up 

The movie [Moneyball](http://en.wikipedia.org/wiki/Moneyball_(film)) focuses on
the "quest for the secret of success in baseball". It follows a low-budget team, 
the Oakland Athletics, who believed that underused statistics, such as a player's 
ability to get on base, better predict the ability to score runs than typical 
statistics like home runs, RBIs (runs batted in), and batting average. Obtaining 
players who excelled in these underused statistics turned out to be much more 
affordable for the team.

In this lab we'll be looking at data from all 30 Major League Baseball teams and
examining the linear relationship between runs scored in a season and a number 
of other player statistics. Our aim will be to summarize these relationships 
both graphically and numerically in order to find which variable, if any, helps 
us best predict a team's runs scored in a season.

## The data

Let's load up the data for the 2011 season.

```{r load-data, eval=TRUE}
load("more/mlb11.RData")
library(ggplot2)
```


In addition to runs scored, there are seven traditionally used variables in the 
data set: at-bats, hits, home runs, batting average, strikeouts, stolen bases, 
and wins. There are also three newer variables: on-base percentage, slugging 
percentage, and on-base plus slugging. For the first portion of the analysis 
we'll consider the seven traditional variables. At the end of the lab, you'll 
work with the newer variables on your own.

1.  What type of plot would you use to display the relationship between `runs` 
    and one of the other numerical variables? Plot this relationship using the 
    variable `at_bats` as the predictor. Does the relationship look linear? If 
    you knew a team's `at_bats`, would you be comfortable using a linear model 
    to predict the number of runs?
    
**In looking at the relationship between two numeric variables, we should use a scatter plot.**
    
```{r}
ggplot(mlb11) + 
  aes(x = at_bats, y = runs) + 
  geom_point() + 
  labs(title = "Runs v at Bats")

```

**There is a positive linear correlation between runs and at_bats. The relationship looks to be semi to moderately strong. I would feel comfortable using a linear model to predict number of runs if I was looking for an average estimate. This relationship isn't incredibly strong, so I don't think the model would be super accurate, but it would probably be directional.**    

If the relationship looks linear, we can quantify the strength of the
relationship with the correlation coefficient.

```{r cor, eval=TRUE}
cor(mlb11$runs, mlb11$at_bats)
```

## Sum of squared residuals

Think back to the way that we described the distribution of a single variable. 
Recall that we discussed characteristics such as center, spread, and shape. It's
also useful to be able to describe the relationship of two numerical variables, 
such as `runs` and `at_bats` above.

2.  Looking at your plot from the previous exercise, describe the relationship 
    between these two variables. Make sure to discuss the form, direction, and 
    strength of the relationship as well as any unusual observations.
    
**As I mentioned above, there is a positive linear correlation between runs and at_bats, meaning the trend is moving up and to the right. The relationship looks to be semi to moderately strong - meaning the data points are not clustered tightly along a linear line, however, they are not spread so far that we can't determine the relationship. There is definitely enough strength to the relationship to tell that these variables are positively correlated. There do appear to be several outliers on the far right hand side of the chart, although, they are not what we would term as "extreme" outliers.**    

Just as we used the mean and standard deviation to summarize a single variable, 
we can summarize the relationship between these two variables by finding the 
line that best follows their association. Use the following interactive 
function to select the line that you think does the best job of going through 
the cloud of points.

```{r plotss-atbats-runs, eval=TRUE}
plot_ss(x = mlb11$at_bats, y = mlb11$runs)
```

After running this command, you'll be prompted to click two points on the plot 
to define a line. Once you've done that, the line you specified will be shown in
black and the residuals in blue. Note that there are 30 residuals, one for each 
of the 30 observations. Recall that the residuals are the difference between the
observed values and the values predicted by the line:

\[
  e_i = y_i - \hat{y}_i
\]

The most common way to do linear regression is to select the line that minimizes
the sum of squared residuals. To visualize the squared residuals, you can rerun 
the plot command and add the argument `showSquares = TRUE`.

```{r plots-atbats-runs-squares, eval=TRUE}
plot_ss(x = mlb11$at_bats, y = mlb11$runs, showSquares = TRUE)
```

Note that the output from the `plot_ss` function provides you with the slope and
intercept of your line as well as the sum of squares.

3.  Using `plot_ss`, choose a line that does a good job of minimizing the sum of
    squares. Run the function several times. What was the smallest sum of 
    squares that you got? How does it compare to your neighbors?
    
**I ran the code an additional four times and the smallest sum of squares I got was 12,6473.**
    
```{r plotss-atbats-runs-squares, eval=TRUE}

plot_ss(x = mlb11$at_bats, y = mlb11$runs, showSquares = TRUE)

```

## The linear model

It is rather cumbersome to try to get the correct least squares line, i.e. the 
line that minimizes the sum of squared residuals, through trial and error. 
Instead we can use the `lm` function in R to fit the linear model (a.k.a. 
regression line).

```{r m1, eval=TRUE}
m1 <- lm(runs ~ at_bats, data = mlb11)
```

The first argument in the function `lm` is a formula that takes the form 
`y ~ x`. Here it can be read that we want to make a linear model of `runs` as a 
function of `at_bats`. The second argument specifies that R should look in the 
`mlb11` data frame to find the `runs` and `at_bats` variables.

The output of `lm` is an object that contains all of the information we need 
about the linear model that was just fit. We can access this information using 
the summary function.

```{r summary-m1, eval=TRUE}
summary(m1)
```

Let's consider this output piece by piece. First, the formula used to describe 
the model is shown at the top. After the formula you find the five-number 
summary of the residuals. The "Coefficients" table shown next is key; its first 
column displays the linear model's y-intercept and the coefficient of `at_bats`.
With this table, we can write down the least squares regression line for the 
linear model:

\[
  \hat{y} = -2789.2429 + 0.6305 * atbats
\]

One last piece of information we will discuss from the summary output is the 
Multiple R-squared, or more simply, $R^2$. The $R^2$ value represents the 
proportion of variability in the response variable that is explained by the 
explanatory variable. For this model, 37.3% of the variability in runs is 
explained by at-bats.

4.  Fit a new model that uses `homeruns` to predict `runs`. Using the estimates 
    from the R output, write the equation of the regression line. What does the 
    slope tell us in the context of the relationship between success of a team 
    and its home runs?
    
```{r}
hr <- lm( runs ~ homeruns, data = mlb11 )
summary(hr)
```
 
**The equation for the regression line is y = 415.2389 + 1.8345 * homeruns. The slope is positive, so we know that the homeruns have a positive relationship with runs. For every homerun made, there are 1.8345 runs made on average.**    

## Prediction and prediction errors

Let's create a scatterplot with the least squares line laid on top.

```{r reg-with-line, eval=TRUE}
plot(mlb11$runs ~ mlb11$at_bats)
abline(m1)
```

The function `abline` plots a line based on its slope and intercept. Here, we 
used a shortcut by providing the model `m1`, which contains both parameter 
estimates. This line can be used to predict $y$ at any value of $x$. When 
predictions are made for values of $x$ that are beyond the range of the observed
data, it is referred to as *extrapolation* and is not usually recommended. 
However, predictions made within the range of the data are more reliable. 
They're also used to compute the residuals.

5.  If a team manager saw the least squares regression line and not the actual 
    data, how many runs would he or she predict for a team with 5,578 at-bats? 
    Is this an overestimate or an underestimate, and by how much? In other 
    words, what is the residual for this prediction?
    
**First lets use our equation from above to calculate where the point would fall and then we'll plot it on our chart below:**

```{r}
#y = -2789.2429 + 0.6305 * atbats

y = -2789.2429 + 0.6305 * 5578
y

```

**If a manager used the least squares regression line to predict runs for a team with 5,578 at-bats, she would predict ~728 runs. In looking at the chart below (red trianlge is prediction), it looks like they would be slightly overestimating the amount of runs. **

```{r}
##with ggplot
ggplot(data = mlb11) + 
  aes(x = at_bats, y = runs) + 
  geom_point() + 
  geom_smooth( method = "lm", se = TRUE, level = 0.95) + 
  geom_point(aes(x = 5578, y = 727.6861), color = "red", shape = 17, size = 3)#se shows confidence interval and level sets it. 0.95 is the default
```

To calculate the residual we will find the closes point and calculate it's difference from our calculated value of runs
```{r}
mlb11[mlb11$at_bats >= 5570 & mlb11$at_bats <= 5580, ]

```    

```{r}
#residual formula is observed - predicted
observed <- 713
predicted <- y 
residual <- observed - predicted
residual
```

## Model diagnostics

To assess whether the linear model is reliable, we need to check for (1) 
linearity, (2) nearly normal residuals, and (3) constant variability.

*Linearity*: You already checked if the relationship between runs and at-bats
is linear using a scatterplot. We should also verify this condition with a plot 
of the residuals vs. at-bats. Recall that any code following a *#* is intended
to be a comment that helps understand the code but is ignored by R.

```{r residuals, eval=TRUE}
plot(m1$residuals ~ mlb11$at_bats)
abline(h = 0, lty = 3)  # adds a horizontal dashed line at y = 0
```

6.  Is there any apparent pattern in the residuals plot? What does this indicate
    about the linearity of the relationship between runs and at-bats?
    
**In looking at the above residuals plot, it looks like the plot's dispersion is fairly random with an equal amount of points both above and below the line. This indicates to us that there is indeed a linear relationship between at_bats and runs.**


*Nearly normal residuals*: To check this condition, we can look at a histogram

```{r hist-res, eval=TRUE}
hist(m1$residuals)
```

or a normal probability plot of the residuals.

```{r qq-res, eval=TRUE}
qqnorm(m1$residuals)
qqline(m1$residuals)  # adds diagonal line to the normal prob plot
```

7.  Based on the histogram and the normal probability plot, does the nearly 
    normal residuals condition appear to be met?
    
**I have some slight hesitation here, but as we are looking for nearly normal, I think we can say that this condition has been met. The histogram does follow a fairly normal shape except for the bucket between 0 and 50. In addition, looking at the Q-Q plot, the points are hovering around the diagonal line fairly closely, although not perfectly. **

*Constant variability*:

8.  Based on the plot in (1), does the constant variability condition appear to 
    be met?
    
**Based on plot 1, it does look like the condition for constant variablity has been met. In looking at the above residuals plot, it looks like the plot's dispersion is fairly random with an equal amount of points both above and below the line. **
    
* * *

## On Your Own

-   Choose another traditional variable from `mlb11` that you think might be a 
    good predictor of `runs`. Produce a scatterplot of the two variables and fit 
    a linear model. At a glance, does there seem to be a linear relationship?

```{r}
ggplot(mlb11) + 
  aes(x = bat_avg, y = runs) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE)
```

**At a glance, this relationship does look to be a positive linear relationship. It also looks to be a stronger relationship than at_bats as well.**
    

-   How does this relationship compare to the relationship between `runs` and 
    `at_bats`? Use the R$^2$ values from the two model summaries to compare. 
    Does your variable seem to predict `runs` better than `at_bats`? How can you
    tell?
    
    ```{r}
batting_avg <- lm( runs ~ bat_avg, data = mlb11 )
summary(batting_avg)
```

**It appears that the relationship between bat_avg and runs is stronger (i.e. a better predictor of runs) than the relationship between at_bats and runs. Looking at the r^2 value of the at_bats model, we got a value of 37.3%, which means that 37.3% of variability in runs is explained by at_bats. When we look at the r^2 value for bat_avg, my r^2 value is 65.61%. Here we can say that 65.61% of the variability in runs has been explained by bat_avg. Based on this, it appears that bat_avg predicts runs better than at_bats. Also, if you look at the scatter plot for bat_avg vs runs, you can see the points are tighter and follow a tighter positive linear trend than does at_bats.**

-   Now that you can summarize the linear relationship between two variables, 
    investigate the relationships between `runs` and each of the other five 
    traditional variables. Which variable best predicts `runs`? Support your 
    conclusion using the graphical and numerical methods we've discussed (for 
    the sake of conciseness, only include output for the best variable, not all 
    five).
    
**Lucky guess, but it looks like the variable I chose before, batting average is the best predictor of runs.**   

```{r}
ggplot(mlb11) + 
  aes(x = bat_avg, y = runs) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE)
```
    
```{r}
batting_avg <- lm( runs ~ bat_avg, data = mlb11 )
summary(batting_avg)
```
  
**Here our r^2 value is 65.61%, which is higher than any of the other 6 traditional variables. This means that we can explain 65.61% of the variation in runs with bat_avg. In looking at the scatter plot above, you can see that the relationship is fairly strong and linear. Most of the points fall fairly close to the least squares line. **    

-   Now examine the three newer variables. These are the statistics used by the 
    author of *Moneyball* to predict a teams success. In general, are they more 
    or less effective at predicting runs that the old variables? Explain using 
    appropriate graphical and numerical evidence. Of all ten variables we've 
    analyzed, which seems to be the best predictor of `runs`? Using the limited 
    (or not so limited) information you know about these baseball statistics, 
    does your result make sense?
 
**We'll first look at on base percentage:** 
    
```{r}
ggplot(mlb11) + 
  aes(x = new_onbase, y = runs) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE)
```

**Without seeing the numerical output, on base percentage looks to be the strongest relationship with runs we have seen so far. Now let's take a look at the numbers:**
    
```{r}
new_onb <- lm( runs ~ new_onbase, data = mlb11 )
summary(new_onb)
```
    
**Wow! The r^2 value is significantly higher than any of the other variables we have seen so far. r^2 value is 84.91%.**

**Now lets take a look at slugging percentage:**

```{r}
ggplot(mlb11) + 
  aes(x = new_slug, y = runs) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE)
```

```{r}
new_slg <- lm( runs ~ new_slug, data = mlb11 )
summary(new_slg)
```

**The r^2 value is even higher than the onbase percentage. r^2 value is 89.69%.**

**Now lets take a look at the onbase + slugging percentage:**

```{r}
ggplot(mlb11) + 
  aes(x = new_obs, y = runs) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE)
```

```{r}
new_obs <- lm( runs ~ new_obs, data = mlb11 )
summary(new_obs)
```

**Incredible! It looks like onbase + slugging percentage is the highest predictor or runs. The r^2 value here is 93.49%, meaning that we can explain 93.49% of the variability in runs with this single variable. This does make sense. Slugging percentage is similar to batting average except it weights each hit higher for each base they get to from a hit. This means that teams with higher slugging percentages are getting more bases (and in turn more runs) for each hit.This in combination with On Base Percentage, which is essentially the % of time they make it to a base for each time to bat, is a powerful predictor. If you have a high on base percentage along with a high slugging percentage, you are not only getting on base, but you are making it to more bases in less times at the plate.**

-   Check the model diagnostics for the regression model with the variable you 
    decided was the best predictor for runs.
    
**Let's check the model diagnostics for the regression model for new_obs (on base percentage + slugging percentage). We will check three things:**

* Linearity - We can tell by looking at the scatter plot above that the relationship is linear.In addition we can look at the residuals and confirm that there is random variation and an equal amount of points above and below the line. In seeing no trends/patterns, we can be confident that this relationship is linear.  


```{r}
ggplot(lm( runs ~ new_obs, data = mlb11 )) + 
  aes(x =.fitted, y = .resid) + 
  geom_point() + 
  geom_hline( yintercept = 0)
```


* Nearly normal residuals

We can look at a histogram to see the distribution of the residuals**

```{r}
hist(new_obs$residuals)
```

**The histogram above looks nearly normal, which gives us confidence that this is a linear relationship. We will also check a Q-Q plot as well:**

```{r}
qqnorm(new_obs$residuals)
qqline(new_obs$residuals)
```

**Looking at the above Q-Q plot, it provides further evidence that this is a linear relationship.**

**The last condition we will check is if there is constant variability. Looking again at our residual plot above, we can see that there is fairly constant and random variability. **


<!-- This is a product of OpenIntro that is released under a Creative Commons Attribution-ShareAlike 3.0 Unported. This lab was written by Mine Çetinkaya-Rundel and Andrew Bray. -->
